---
title: "Multi-Path AIC: Diabetes Progression"
author: "Han Xiu, Lauren Cooper, Fritzner Pierre"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diabetes progression (multi-path AIC)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# 1. Setup

We begin by loading the required packages.\
The functions `build_paths()`, `stability()`, and `plausible_models()` come from the package developed for this project.

The diabetes progression dataset (`efron2004`) is available in the **care** package, which we load as well.

```{r setup, message = FALSE}
library(modelselection13)
library(care)

set.seed(28)
```

# 2. Data preparation

We use the diabetes progression dataset from Efron et al. (2004), available in the `care` package.\
The dataset contains 442 patients with 10 baseline predictors (age, sex, BMI, mean arterial pressure, and six serum measures).

To illustrate model complexity and allow the multi-path search to explore richer model spaces,\
we follow the assignment instructions and augment the baseline predictors with second-order terms:\
- quadratic terms for each baseline variable,\
- all pairwise interactions.

This produces approximately 60 predictors in total.\
We then create a 70/30 train–test split that will be used for model selection and evaluation throughout the vignette.

```{r data-prep}
# Load the diabetes progression data from the 'care' package
data("efron2004", package = "care")

# Convert to a plain numeric matrix (remove AsIs)
X_base_mat <- as.matrix(unclass(efron2004$x))
dim(X_base_mat) 

# Convert to data.frame
X_base <- as.data.frame(X_base_mat)
dim(X_base) 

y <- as.numeric(efron2004$y[, 1])

# ---- Build second-order terms: quadratics + pairwise interactions ----

# Quadratic terms: x_j^2 for each baseline predictor
X_quad <- X_base^2
colnames(X_quad) <- paste0(colnames(X_base), "_sq")

# Pairwise interactions: x_j * x_k for j < k
p <- ncol(X_base)
interaction_list <- vector("list", length = p * (p - 1) / 2)
interaction_names <- character(length(interaction_list))

idx <- 1L
for (j in seq_len(p - 1L)) {
  for (k in (j + 1L):p) {
    interaction_list[[idx]] <- X_base[[j]] * X_base[[k]]
    interaction_names[idx] <- paste0(colnames(X_base)[j], ":", colnames(X_base)[k])
    idx <- idx + 1L
  }
}
X_int <- as.data.frame(interaction_list)
colnames(X_int) <- interaction_names

# Combine all predictors: main effects + quadratics + interactions
X_full <- cbind(X_base, X_quad, X_int)

dim(X_full)

# ---- Train/test split (e.g., 70/30) ----

n <- nrow(X_full)
train_prop <- 0.7
n_train <- floor(train_prop * n)

set.seed(2025)  # reproducible split
train_idx <- sample(seq_len(n), size = n_train, replace = FALSE)

X_train <- X_full[train_idx, , drop = FALSE]
y_train <- y[train_idx]

X_test  <- X_full[-train_idx, , drop = FALSE]
y_test  <- y[-train_idx]

# Quick sanity checks
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

# 3. Multi-path forward selection using AIC

To explore the model space efficiently, we apply our **multi-path forward selection algorithm** to the training data.\
Unlike traditional forward selection, which commits to a single best variable at each step, multi-path search allows **near-optimal alternatives** (within a small AIC tolerance) to remain active.\
This produces a **forest** of plausible model trajectories, capturing competing explanations that fit the data almost equally well.

Using the training predictors $X_{\text{train}}$ and response $y_{\text{train}}$, we run `build_paths()` with:

-   $K = 10$: maximum model size\
-   $\varepsilon = 10^{-6}$: minimum AIC improvement required to expand a model\
-   $\Delta = 2$: AIC near-tie tolerance\
-   $L = 30$: maximum number of models retained per step

The result is a structured collection of models at each forward-selection step, which we will later analyze for stability.

```{r build-paths, message=FALSE}
# Multi-path AIC forward selection on the training set
set.seed(28)

K_val <- 10     # max number of variables to include
eps_val <- 1e-6 # minimum AIC improvement
delta_val <- 2  # allow small AIC near-ties
L_val <- 30     # limit models per step

paths <- build_paths(
  x = X_train,
  y = y_train,
  family = "gaussian",
  K = K_val,
  eps = eps_val,
  delta = delta_val,
  L = L_val
)

paths$meta
length(paths$frontiers)          # number of steps+1
paths$frontiers[[1]]             # step 0: empty model
paths$frontiers[[2]][1:5, ]      # first few models at step 1
```

# 4. Stability analysis via resampling

To assess how reliably each predictor enters the model across perturbed versions of the data, we perform a **bootstrap-based stability analysis**.

For each bootstrap resample of the training set:

1.  `build_paths()` is rerun on the resampled data.\
2.  For every predictor $j$, we record the proportion of discovered models in which $j$ appears.\
3.  These proportions are averaged over $B = 50$ resamples to obtain a stability score\
    $\pi_j \in [0, 1]$.

Predictors with high $\pi_j$ are consistently selected across resamplings, while those with low stability tend to enter models only sporadically.\
The stability vector $\pi$ will guide our selection of a final set of plausible models.

```{r stability, message=FALSE}
set.seed(28)

stab <- stability(
  x = X_train,
  y = y_train,
  B = 50,
  resample = "bootstrap",
  family = "gaussian",
  K = K_val,     
  eps = eps_val,
  delta = delta_val,
  L = L_val
)

# Show first few stability scores
head(stab$pi)
summary(stab$pi)
```

# 5. Constructing the plausible model set

To balance statistical fit and model robustness, we combine the AIC results from the multi-path search with the stability scores $\pi_j$ obtained from the bootstrap analysis.

We select two tuning parameters:

-   **AIC tolerance** $\Delta = 2$:\
    We use an AIC tolerance of $\Delta = 2$, since models within 2 AIC units are generally considered statistically indistinguishable.

-   **Stability threshold** $\tau = 0.4$:\
    We set the stability threshold to $\tau = 0.4$, retaining only models built from predictors that appear consistently across bootstrap resamples.

We optionally remove near-duplicate models using Jaccard similarity to avoid redundancy.

The resulting plausible model set provides a principled balance between predictive fit and structural stability.

```{r plausible-models}
Delta_val <- 2        # AIC window
tau_val <- 0.4       # stability threshold

plaus <- plausible_models(
  forest = paths,
  pi = stab$pi,
  Delta = Delta_val,
  tau = tau_val,
  jaccard_threshold = 0.8
)

plaus
```
